\section{Introduction}\sectionlabel{introduction}

Naming is a pervasive concern in the design and implementation of
programming languages. Names identify \emph{declarations} of program
entities (variables, functions, types, modules, etc.) and allow these
entities to be \emph{referenced} from other parts of the program. Name
\emph{resolution} associates each reference to its intended
declaration(s), according to the semantics of the language. Name
resolution underlies most operations on languages and programs,
including static checking, translation, mechanized description of
semantics, and provision of editor services in IDEs. Resolution is often
complicated, because it cuts across the local inductive structure of
programs (as described by an abstract syntax tree). For example, the
name introduced by a \pcfmcode{let} node in an ML AST may be referenced by an
arbitrarily distant child node. Languages with explicit name spaces
lead to further complexity; for example, resolving a qualified
reference in Java requires first resolving the class or package name to
a context, and then resolving the member name within that context. 
But despite this diversity, 
it is intuitively clear that the basic concepts of resolution reappear in
similar form across a broad range of lexically-scoped languages.

In practice, the name resolution rules of real programming
languages are usually described using {\it ad hoc} and informal
mechanisms. Even when a language \emph{is} formalized, its resolution
rules are typically encoded as part of static and dynamic judgments
tailored to the particular language, rather than being presented
separately using a uniform mechanism. This lack of modularity in
language description is mirrored in the implementation of language
tools, where the resolution rules are often encoded multiple times to
serve different purposes, e.g., as the manipulation of a symbol table in
a compiler, a use-to-definition display in an IDE, or a substitution
function in a mechanized soundness proof. This repetition results in
duplication of effort and risks inconsistencies. To see how much better
this situation might be, we need only contrast it with 
the realm of syntax definition, where context-free grammars provide 
a well-established declarative formalism that underpins a 
wide variety of useful tools. 

\paragraph{Formalizing resolution.}\label{formalizing-resolution}
This paper describes a formalism that we believe can help play 
a similar role for name resolution in lexically-scoped languages. 
It consists of a
\emph{scope graph}, which represents the naming structure of a program,
and a \emph{resolution calculus}, which describes how to resolve
references to declarations within a scope graph. The scope graph
abstracts away from the details of a program AST, leaving just the
information relevant to name resolution. Its nodes include name
references, declarations, and ``scopes,'' which (in a slight abuse of
conventional terminology) we use to mean minimal program regions that
behave uniformly with respect to name resolution. 
Edges in the scope graph associate
references to scopes, declarations to scopes, or scopes to ``parent''
scopes (corresponding to lexical nesting in the original program AST).
The resolution calculus specifies how to construct a path through
the graph from a reference to a declaration, which corresponds to 
a possible resolution of the reference. Hiding of one definition
by a ``closer'' definition is modeled by providing
an ordering on resolution paths. Ambiguous references 
correspond naturally to multiple resolution paths starting
from the same reference node; unresolved references correspond
to the absence of resolution paths.  
To describe programs involving explicit name
spaces, the scope graph also supports giving names to scopes, and can
include ``import'' edges to make the contents of a named scope visible
inside another scope. The calculus supports complex import patterns including
transitive and cyclic import of scopes.

This language-independent formalism gives us clear, abstract definitions
for concepts such as scope, resolution, hiding, and import. 
We build on these concepts to define generic notions 
of $\alpha$-equivalence and valid renaming.  
We also give a practical algorithm for computing conventional
static environments mapping bound identifiers to the 
AST locations of the corresponding declarations,
which can be used to implement
a deterministic, terminating resolution function that is consistent 
with the calculus.  We expect that the formalism can be used 
as the basis for other language-independent tools. 
In particular, any tool that relies
on use-to-definition information, such as an IDE offering code completion for
identifiers, or a live variable analysis in a compiler, should
be specifiable using scope graphs.

On the other hand, 
the construction of a scope graph from a given program is a
language-\emph{dependent} process. For any given language, the
construction can be specified by a conventional syntax-directed
definition over the language grammar; 
we illustrate this approach for a small language in this paper.  
We would also like a more
generic \emph{binding specification language} which could be
used to describe how to 
construct the scope graph for an arbitrary object language.
We do not present such a language in this paper. However, the
work described here was inspired in part by our previous work on 
NaBL~\cite{KonatKWV12}, a DSL that provides high-level,
non-algorithmic descriptions of name binding and scoping rules 
suitable for use by a  (relatively) naive language designer.
The NaBL implementation integrated into the Spoofax Language Workbench
\cite{KatsV10} automatically generates an incremental name resolution
algorithm that supports services such as code completion and static
analysis. However, the NaBL language itself is defined largely by
example and lacks a high-level semantic description; one might say that
it works well in practice, but not in theory. Because they are
language-independent, scope graphs can be used to give a formal
semantics for NaBL specifications, although we defer detailed
exploration of this connection to further work. 

\paragraph{Relationship to Related Work.}

The study of name binding has received a great deal of attention,
focused in particular on two topics.
The first is how to represent 
(already resolved) programs in a way that makes the binding
structure explicit and supports 
convenient program manipulation ``modulo
$\alpha$-equivalence''~\cite{deBruijn72,DBLP:conf/pldi/PfenningE88,Chargueraud12,GabbayP02,Cheney05a}.
Compared to this work, our system is novel in several significant respects.
(i)~Our representation of program binding structure is {\it independent}
of the underlying language grammar and program AST, with the
benefits described above.
(ii)~We support representation of ill-formed programs, 
in particular, programs with ambiguous or undefined references; such programs
are the normal case in IDEs and other front-end tools. 
(iii)~We support description of binding in languages with
explicit name spaces, such as modules or OO classes, which
are common in practice.  

A second well-studied topic is binding specification languages, 
which are usually enriched grammar descriptions that permit simultaneous 
specification of language syntax and binding
structure~\cite{SewellNOPRSS10,DybvigHB92,HermanW08,StansiferW14,WeirichYS11}.
This work is essentially complementary to the design we present here. 

\paragraph{Specific contributions.}\label{contributions}

\begin{itemize}
  
 \item \emph{Scope Graph and Resolution Calculus}: We introduce a
 language-independent framework to capture the relations among
 \emph{references}, \emph{declarations}, \emph{scopes}, and  \emph{imports} in a
 program. 
We give a declarative specification of the resolution of
 references to declarations by means of a calculus that defines resolution paths
 in a scope graph (\Section{rescalc}).

 \item \emph{Variants}: We illustrate the modularity of our core framework design
by describing several variants that support more complex binding 
schemes (\Section{extensions}).

 \item \emph{Coverage}: We show that the framework covers interesting  name
 binding patterns in existing languages, including various flavors of let bindings,
 qualified names, and inheritance in Java (\Section{coverage}).
  
 \item \emph{Scope graph construction:} We show how scope graphs can be constructed
 for arbitrary programs in a simple example language via straightforward syntax-directed traversal (\Section{construction}).

 \item \emph{Resolution algorithm}: We define a deterministic and terminating
 resolution algorithm based on the construction of binding environments, and
 prove that it is sound and complete with respect to the calculus
 (\Section{resalg}).
 
 \item \emph{$\alpha$-equivalence and renaming}: We define a language-independent
 characterization of $\alpha$-equivalence of programs, and use it to define a 
 notion of valid renaming (\Section{applications}).

\end{itemize}

\techrep{The extended version of this paper \cite{TUD-SERG-2015-001-local}}{The appendixes of this paper} present\techrep{s}{} the encoding 
of additional name binding patterns and the details of the correctness proof 
of the resolution algorithm.


\endinput


\TODO{This is an attempt at relating scope graphs to traditional notions of
scope. Not clear that this is useful (too meta) or where it should go. }

The traditional understanding of `Scope'\footnote{We will use Scope with a
capital S for the traditional sense of scope.} in programming languages is
described in
Wikipedia\footnote{\url{http://en.wikipedia.org/wiki/Scope_(computer_science)}}
as ``[T]he [S]cope of a name binding – an association of a name to an entity,
such as a variable – is the part of a computer program where the binding is
valid: where the name can be used to refer to the entity.'' Thus, a Scope in
this sense is a collection of program points (AST nodes), in which a
declaration can be referred to (in which it is visible).
Such a notion of Scope is the \emph{outcome} of name resolution and cannot be
used as a \emph{building block} for a \emph{theory of name resolution}.
Each language and each scoping construct can define a complicated function,
selecting a more or less arbitrary set of AST nodes to compose the Scope for a
declaration, inhibiting a general, language independent theory of name
resolution.
Even if a Scope assignment is more well behaved, it is non-modular.
Two declarations defined close to each other, say in the same block, do not
necessarily have the same Scope, since the Scope of a declaration may be
affected by \emph{other} declarations (for the same name) through shadowing.

To remedy this situation we introduce a theory that provides a principled
procedure for determining the Scope of declarations.
The basic building block of our theory is a \emph{scope} (with lower case s;
for lack of a better term), which corresponds to a \kw{letrec} block (or a module), i.e.
a group of mutually recursive declarations, \emph{not} including the set of all
program points in which these bindings are valid.
We model visibility of declarations in other scopes than the scope in which they
are declared, through \emph{edges} in the scope graph.
Scope graph edges provide a unified model of different scoping regimes.
We first consider \emph{lexical scope}, in which names are declared in the local
lexical environment.
In \Section{imports} we consider \emph{imports}, which make names from a
non-local context available.


