	\maketitle
	
\vspace*{\fill}

This book is on \BookOnGithub{github} at \url{\GithubURL}

\vspace*{\fill}

copyright \copyright{} 2015 \href{http://eelcovisser.org}{Eelco Visser}
	
\newpage
\tableofcontents

\newpage
\chapter*{Preface}

This is a book about declarative language definition with the Spoofax Language
Workbench.

This is a book in progress as Spoofax itself is very much in progress.
Spoofax is the product of an ongoing research project that investigates the
nature of software language definition.
We are continuously experimenting with better ways to do things. 
As a result, new features are being added to Spoofax and we are also always
working on the guts of the workbench.
Therefore, working with Spoofax can be a somewhat rocky experience.

This book and the accompanying Spoofax projects are on \BookOnGithub{github}. 
I very much would appreciate pull requests.

Eelco Visser

\today

\newpage
\chapter{Declarative Language Definition}


\section{What is a Language?}

internal, external, embedded languages

\newpage
\section{Language Workbenches}

what is a language workbench

Xtext, MPS, Rascal, Spoofax

\newpage
\input{getting-spoofax/getting-spoofax}


\newpage
\chapter{Syntax}

\begin{verbatim}
context-free syntax
  Exp.Var = ID
  Exp.Mul = [[Exp] * [Exp]] {left}
  Exp.Add = [[Exp] + [Exp]] {left}
context-free priorities
  Exp.Mul > Exp.Add
\end{verbatim}


\newpage
\chapter{Names \& Types}

\newpage
\chapter{Transformation}

\newpage
\newpage
\chapter{Semantics}

